---
title: "Has Our Electricity Usage Changed Since June 2024?"
date: "`r Sys.Date()`"
output: 
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r include=FALSE, echo=FALSE}
library(tidyverse)
library(lubridate)
library(gt)
library(scales)
# Load your cleaned data
daily_usage <- read_csv("/home/rstudio/output/daily_usage.csv")
t_test_results <- read_csv("/home/rstudio/output/t_test_results.csv")
model_summary <- read_csv("/home/rstudio/output/model_summary.csv")
```

# Data Overview

- Time range: Oct 2023 – July 2025
- Frequency: Half-hour readings aggregated to daily totals
- Metric: kWh used per day

---
# Initial Insight (Raw Comparison)

```{r, echo=FALSE, fig.width=15}
ggplot(daily_usage %>% 
         filter(!Month %in% c("June", "July", "August", "September")),
       aes(x = fct_rev(DayOfWeekShort), y = Daily_kWh, fill = Period)) +
  geom_boxplot() +
  facet_wrap(~Month) +
  labs(title = "Electricity Usage by Weekday (Unadjusted)",
       x = "Day of Week", y = "Daily Usage (kWh)") +
  theme_minimal() +
  coord_flip()
```

---
# Seasonality Observed

```{r, echo=FALSE, fig.width=7}
daily_usage %>% 
  group_by(Month) %>% 
  summarise(mean_kWh = mean(Daily_kWh)) %>% 
  ggplot(aes(x = Month, y = mean_kWh)) +
  geom_col(fill = "steelblue") +
  labs(title = "Average Daily Usage by Month",
       x = "Month", y = "Mean Daily kWh") +
  theme_minimal()

ggplot(daily_usage, aes(x = Date, y = Daily_kWh)) +
  geom_line(alpha = 0.3) +
  geom_smooth(method = "loess", span = 0.2, se = TRUE, color = "blue") +
  geom_vline(xintercept = as.Date("2024-06-01"), linetype = "dashed", color = "red") +
  labs(title = "Daily Electricity Usage Over Time",
       subtitle = "LOESS smoothed trend with June 1, 2024 cutoff",
       y = "Daily kWh", x = "Date") +
  theme_minimal()
```


---
# Adjusting for Seasonality: Linear Model

```{r, echo=FALSE}
model_summary %>%
  filter(term == "PeriodAfter June 2024") %>%
  gt() %>%
  fmt_number(columns = everything(), decimals = 3) %>%
  tab_header(
    title = "Effect of Period After June 2024 (Adjusted for Seasonality)"
  )

daily_usage <- daily_usage %>%
  mutate(
    Period = factor(Period, levels = c("Before June 2024", "After June 2024")),
    DayOfWeek = factor(DayOfWeek, 
                       levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
  )

# Fit the model (if not already done)
model_by_day <- lm(Daily_kWh ~ Period * DayOfWeek + Month, data = daily_usage)

# Get tidy summary with confidence intervals
model_by_day_summary <- broom::tidy(model_by_day, conf.int = TRUE)

# Filter relevant terms: main Period effect (Monday) + Period:DayOfWeek interactions
did_terms <- model_by_day_summary %>%
  filter(str_detect(term, "PeriodAfter June 2024:DayOfWeek") | term == "PeriodAfter June 2024") %>%
  mutate(
    DayOfWeek = case_when(
      term == "PeriodAfter June 2024" ~ "Monday",
      TRUE ~ str_remove(term, "PeriodAfter June 2024:DayOfWeek")
    )
  )

# Extract Monday’s row (main Period effect)
monday_row <- did_terms %>% filter(DayOfWeek == "Monday")

# Calculate absolute estimates for each day by adding Monday's estimate to other days
did_terms_abs <- did_terms %>%
  mutate(
    estimate = if_else(DayOfWeek == "Monday", estimate, estimate + monday_row$estimate),
    term = DayOfWeek  # replace term with day names for clarity
  ) %>%
  select(term, estimate, std.error, statistic, p.value, conf.low, conf.high) %>%
  arrange(factor(term, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")))

# Format and display as a gt table
did_terms_abs %>%
  gt() %>%
  fmt_number(
    columns = vars(estimate, std.error, statistic, p.value, conf.low, conf.high),
    decimals = 3
  ) %>%
  tab_header(
    title = "Estimated Energy Change After June 2024 by Weekday"
  ) %>%
  cols_label(
    term = "Day of Week",
    estimate = "estimate",
    std.error = "std.error",
    statistic = "statistic",
    p.value = "p-value",
    conf.low = "conf.low",
    conf.high = "conf.high"
  )
```

---
# In the above tables:
- Estimate is the change in kWh usage.
- std.error is the amount the estimate varies.
- p-value is the significance. If the p-value is less than 0.05 the estimate is statistically significant.
- We can have 95% confidence that the energy consumption change is between conf.low and conf.high
---
# Difference-in-Differences

```{r, echo=FALSE, fig.width=7}
daily_usage %>% 
  filter(!Month %in% c("June", "July", "August", "September")) %>% 
  mutate(DayOfWeekShort = factor(DayOfWeekShort, 
                                 levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))) %>% 
  group_by(Period, DayOfWeekShort) %>% 
  summarise(mean_kWh = mean(Daily_kWh), .groups = 'drop') %>% 
  ggplot(aes(x = DayOfWeekShort, y = mean_kWh, color = Period, group = Period)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(title = "Daily Usage Before vs. After June 2024",
       x = "Day of Week", y = "Average Daily kWh") +
  theme_minimal()

filtered_data <- daily_usage %>%
  filter(!Month %in% c("June", "July", "August", "September"))

ggplot(filtered_data, aes(x = Period, y = Daily_kWh, fill = Period)) +
  stat_summary(fun = mean, geom = "bar", width = 0.5) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2) +
  labs(
    title = "Adjusted Comparison of Average Daily Usage (Excludes June–Sept)",
    y = "Daily kWh",
    x = "Period"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```


---
# Summary of Findings

- 📈 After adjusting for seasonality, usage increased by **3.2 kWh/day**
- 📊 Saturday showed highest post-June increase
- To accurately say the annual difference in energy consumption we would need more data.
- A quick "back-of-the-napkin" estimate for annual change is a 1,170 kWh increase (3.205 kWh x 365 days).
